---
title: "Using the neutr package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the neutr package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(neutr)
```

The `neutr` package does two things for neutral models of species assemblages:

* **inference**: maximum likelihood estimation of the parameters of the Ewens
  model (`optim.ewens()`), the Etienne model of a local community with dispersal
  limitation (`optim.etienne()`), the Pitman model (`optim.pitman()`) and the
  multideme model (`optim.multideme()`);
* **generation**: fast and exact simulation of typical species abundance
  distributions of the Ewens model (`generate.hoppe.urn0()`,
  `generate.hoppe.urn()`) and of the Pitman model (`generate.pitman.urn()`),
  including for samples of $10^{12}$ individuals and more.

A species abundance distribution is a vector of positive counts
$(n_1, \dots, n_k)$: $k$ species, $J = \sum_i n_i$ individuals.

## Generating a community

Ewens' model has a single parameter $\theta$ (the fundamental biodiversity
number). `generate.hoppe.urn()` returns the abundances, sorted in decreasing
order, together with the *expected* number of species for these parameters:

```{r generate}
set.seed(42)
com <- generate.hoppe.urn(theta = 11.3, J = 234373)
length(com$abundance)   # species in this draw
com$k                   # expected number of species
head(com$abundance)
sum(com$abundance)      # always exactly J
```

```{r rank-abundance, fig.cap = "Rank-abundance curve of the simulated community."}
plot(com$abundance, log = "y", xlab = "Rank", ylab = "Abundance", pch = 16, cex = 0.6)
```

The generator works with the Griffiths-Engen-McCloskey representation of the
model, so its cost hardly depends on the number of individuals:

```{r huge}
huge <- generate.hoppe.urn(theta = 100, J = 1e12)
length(huge$abundance)
sum(huge$abundance) == 1e12
```

`generate.hoppe.urn0()` simulates Hoppe's urn exactly, individual by individual
(vectorised, so still fast for samples of millions). It is mostly useful as a
reference. It draws random numbers with the `dqrng` package, which has its own
random stream: **`set.seed()` does not affect it**, use its `seed` argument:

```{r urn0}
a <- generate.hoppe.urn0(theta = 11.3, J = 1e5, seed = 1)
b <- generate.hoppe.urn0(theta = 11.3, J = 1e5, seed = 1)
identical(a, b)
```

The Pitman model adds a discount parameter $\sigma \in (0, 1)$, which gives
heavier tails (more rare species) than Ewens' model:

```{r pitman-gen}
set.seed(1)
pit <- generate.pitman.urn(theta = 10, sigma = 0.3, J = 1e5)
c(species = length(pit$abundance), expected = round(pit$k, 1))
```

## Estimating the parameters

`optim.ewens()` gives the maximum likelihood estimate of $\theta$, which solves
$k = \theta[\psi(\theta + J) - \psi(\theta)]$:

```{r ewens}
ab <- c(234, 87, 34, 5, 4, 3, 3, 2, 2, 1, 1, 1, 1)
ew <- optim.ewens(ab)
ew
```

`optim.pitman()` estimates $(\theta, \sigma)$:

```{r pitman}
pm <- optim.pitman(ab, init_vals = c(10, 0.1))
pm
```

### Comparing the two models

The Ewens model is the Pitman model with $\sigma = 0$, so the two are nested and
can be compared with a likelihood-ratio test or AIC. The `logl` element returned
by `optim.ewens()` and `optim.pitman()` is on the same scale for both functions:

```{r lrt}
lrt <- 2 * (pm$logl - ew$logl)
c(LRT = lrt, p.value = pchisq(lrt, df = 1, lower.tail = FALSE))
c(Ewens = -2 * ew$logl + 2 * 1, Pitman = -2 * pm$logl + 2 * 2)   # AIC
```

(As $\sigma = 0$ lies on the boundary of the parameter space, the $\chi^2_1$
p-value is conservative; halve it for the usual boundary correction.)

### Recovering known parameters

```{r recover}
set.seed(2024)
sim <- generate.pitman.urn(theta = 10, sigma = 0.3, J = 1e5)$abundance
fit <- optim.pitman(sim)
round(c(theta = fit$theta, sigma = fit$sigma), 3)   # truth: 10 and 0.3
```

### Boundary estimates

When the likelihood has no interior maximum for the data at hand, the estimate
lands on the edge of the search interval and the functions issue a warning:

```{r boundary}
tryCatch(optim.ewens(rep(1, 20)), warning = function(w) conditionMessage(w))
```

## A local community with dispersal limitation: the Etienne model

In Etienne's (2005) model, a local community of `J` individuals receives immigrants
from a neutral metacommunity with fundamental biodiversity number $\theta$, at rate
$m$. `optim.etienne()` estimates $(\theta, m)$ from a single sample; on the example
of Etienne's paper it reproduces his estimates $\theta = 7.05$ and $m = 0.226$:

```{r etienne}
zoo <- c(8, 5, 3, 2, 1, 1)
et <- optim.etienne(zoo)
et
```

The model with $m = 1$ (no dispersal limitation) is the Ewens model, and the `logl`
values of the two functions are on the same scale:

```{r etienne-lrt}
ew_zoo <- optim.ewens(zoo)
lrt <- 2 * (et$logl - ew_zoo$logl)
c(LRT = lrt, p.value = pchisq(lrt, df = 1, lower.tail = FALSE))
```

(As for the Pitman model, $m = 1$ lies on the boundary of the parameter space and
the $\chi^2_1$ p-value is conservative.) When the data show no dispersal limitation,
the estimate of $m$ reaches its upper bound, with a warning, and the Etienne model
reduces to the Ewens model; this is the case for the sample `ab` used above:

```{r etienne-boundary}
suppressWarnings(optim.etienne(ab))[c("theta", "m", "logl")]
ew_ab <- optim.ewens(ab)
ew_ab
```

The likelihood itself is given by `logl.etienne()`. It depends on the data through
the coefficients $K(D,A)$, which are the expensive part of the calculation (their
cost grows as $J^2$): compute them once with `logkda()` to evaluate the likelihood
many times, for instance along a profile:

```{r etienne-profile, fig.cap = "Profile of the log-likelihood in m, at the estimated theta."}
lk <- logkda(zoo)
m <- seq(0.02, 1, by = 0.01)
plot(m, logl.etienne(zoo, et$theta, m, log_kda = lk), type = "l",
     xlab = "immigration rate m", ylab = "log-likelihood")
abline(v = et$m, lty = 2)
```

### Uncertainty and local maxima

`optim.etienne()` returns standard errors (`se_theta`, `se_m`, `se_I`) from the
inverse Hessian, and with `ci = TRUE` likelihood-ratio confidence intervals, which
are preferable because the likelihood is often strongly asymmetric. A limit equal to
a bound of the search interval means that the interval extends at least that far:

```{r etienne-ci}
fit <- optim.etienne(zoo, ci = TRUE)
rbind(theta = c(estimate = fit$theta, se = fit$se_theta, fit$ci_theta),
      m     = c(estimate = fit$m,     se = fit$se_m,     fit$ci_m))
```

The likelihood surface can have several local maxima; all those found are listed in
`local_maxima`, the global one first:

```{r etienne-local}
optim.etienne(c(15, 81, 80, 2, 1, 2))$local_maxima
```

### Data files of the TeTame software

The TeTame software (Chave and Jabot) reads species abundances from a text file,
one value per line, samples being separated by a line containing `&`. Such a file
is easily read in R. Here we write the trial file distributed with TeTame 2.1:

```{r tetame-file}
file <- tempfile(fileext = ".txt")
writeLines(c(1, 8, 80, 2, 1, 20, 0, "&", 15, 8, 8, 20, 10, 2, 0, "&",
             15, 81, 80, 2, 1, 2, 0), file)

x <- scan(file, what = "", quiet = TRUE)
samples <- unname(split(as.numeric(x[x != "&"]), cumsum(x == "&")[x != "&"]))
```

TeTame's option `e` estimates $(\theta, m)$ in each sample. TeTame reports the
negative log-probability of the abundance vector, which is `-logl.etienne(..., full = TRUE)`:

```{r tetame-e}
do.call(rbind, lapply(samples, function(y) {
  fit <- suppressWarnings(optim.etienne(y))
  data.frame(S = length(y[y > 0]), J = sum(y), theta = fit$theta, se_theta = fit$se_theta,
             m = fit$m, se_m = fit$se_m,
             negloglik = -logl.etienne(y, fit$theta, fit$m, full = TRUE),
             n_local_maxima = nrow(fit$local_maxima))
}))
```

TeTame's option `j` (Jabot et al. 2008) estimates $m$ in each sample, the regional
pool being the sum of the samples: this is `optim.multideme()`, with one row per
sample (the species must be in the same order in every sample):

```{r tetame-j}
optim.multideme(do.call(rbind, samples), ci = TRUE)[c("I", "se_I", "m", "se_m", "ci_m")]
```

The likelihood surface is computed on a grid with `logl.etienne()`, which is
vectorised; the contour delimits the joint 95% confidence region:

```{r tetame-surface, fig.cap = "Log-likelihood surface of the second sample of the TeTame trial file; the contour delimits the joint 95% confidence region and the point is the maximum likelihood estimate."}
y <- samples[[2]]
fit <- optim.etienne(y)
lk <- logkda(y)                                  # computed once for the whole grid
grid <- expand.grid(theta = exp(seq(log(0.5), log(200), length.out = 80)),
                    m = seq(0.005, 1, length.out = 80))
grid$logl <- logl.etienne(y, grid$theta, grid$m, log_kda = lk)
z <- matrix(grid$logl - fit$logl, 80, 80)
image(log(unique(grid$theta)), unique(grid$m), pmax(z, -10), col = hcl.colors(50),
      xlab = "log(theta)", ylab = "m")
contour(log(unique(grid$theta)), unique(grid$m), z, levels = -qchisq(0.95, 2) / 2,
        drawlabels = FALSE, add = TRUE)
points(log(fit$theta), fit$m, pch = 19)
```

## The multideme model

`optim.multideme()` takes a matrix with **one row per deme (local community) and
one column per species**. The regional abundance distribution is the sum of the
rows. For each deme it estimates the rescaled immigration rate
$I = \frac{m}{1 - m}(J - 1)$ and the immigration rate $m$:

```{r multideme}
demes <- rbind(
  c(44, 37, 34, 5, 4, 3, 3, 2, 2, 1, 1, 1, 1),
  c(240, 20, 48, 2, 21, 1, 3, 2, 5, 2, 0, 1, 1)
)
optim.multideme(demes)
```

Use `verbose = TRUE` to follow the progress over many demes.

## Expected number of species

`kest()` is the exact expected number of species in the Ewens model.
`kest.gt()` and `kest.gt0()` give the expected number of species with more than
`N` individuals, deterministically (an approximation, via an integral) and by
simulation:

```{r richness}
kest(11.3, 234373)
kest.gt(11.3, 234373, N = 50)$value
set.seed(3)
kest.gt0(11.3, 234373, N = 50, nrep = 50)
```
